---
title: "Air polution"
output: html_notebook
---
```{r}
library(readxl)
library(tidyverse)
library(sf)
library(ggplot2)
library(dplyr)
library(readxl)
```
#reading in the datasets that will be used in this project. 

```{r}
air_pollution_us <- read.csv("data/pollution_us_2000_2016.csv")
```

```{r}
air_pollution_global <- read.csv("data/data.csv")
```

#the following datasets contain information of total population Nox and PM globally between 2003 - 2018

```{r}
population <- read_excel("data/aqdh-country-trends-major-air-pollutants-2003-2018.xlsx", sheet = "Population (GPWv4.11)" )
NOx_Population <- read_excel("data/aqdh-country-trends-major-air-pollutants-2003-2018.xlsx", sheet = "NOx Population-Weighted (ppm)")
PM_Pop <- read_excel("data/aqdh-country-trends-major-air-pollutants-2003-2018.xlsx", sheet = "PM Pop.-Weighted (kg m^-3)")
```

#Droping duplicates in the dataframes. 

```{r}
NOx_Population
```

```{r}
population <- population %>% filter(!duplicated(.))
```

```{r}
NOx_Population  <- NOx_Population %>% filter(!duplicated(.))
```

```{r}
PM_Pop  <- PM_Pop %>% filter(!duplicated(.))
```

#Reshaping the dataframes into a large table

```{r}
population <- population%>%
 pivot_longer(cols = c('2003', '2004', '2005', '2006', '2007', '2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018'),
               names_to = 'year',
               values_to = 'population')

```


```{r}
NOx_Population <- NOx_Population%>%
  pivot_longer(cols = c('2003', '2004', '2005', '2006', '2007', '2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018'),
               names_to = 'year',
               values_to = 'nox')

```

```{r}
PM_Pop <- PM_Pop%>%
  pivot_longer(cols = c('2003', '2004', '2005', '2006', '2007', '2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018'),
               names_to = 'year',
               values_to = 'PM')

```

#Merging the datasets into a large dataframe. 
```{r}
all_population <- population %>%
  full_join(NOx_Population, by = c("country", "year"))%>%
  full_join(PM_Pop, by = c("country", "year"))
all_population
```

#Dropping extra information that is not needed.
```{r}
air_pollution_global
```
#Dropping duplicates 
```{r}
air_pollution_global <- air_pollution_global%>% filter(!duplicated(.))
```

```{r}
air_pollution_filtered_gbl <- subset(air_pollution_global, select = -c(IndicatorCode, ValueType, 
Location.type, Period.type, IsLatestYear, Dim1.type, Dim1ValueCode, Dim2.type, Dim2ValueCode, Dim3.type, Dim3, Dim3ValueCode, DataSourceDimValueCode, DataSource, FactValueNumericPrefix, FactValueUoM, FactValueNumericLowPrefix, FactValueNumericLow, FactValueNumericHighPrefix, FactValueNumericHigh, FactValueTranslationID, FactComments, Language, DateModified))
air_pollution_filtered_gbl
```
#Renaming columns in the Global Dataframe.
```{r}
air_pollution_filtered_gbl <- air_pollution_filtered_gbl%>%
  transmute(
    continent = ParentLocation,
    country = Location,
    year = Period,
    sex = Dim1, 
    cause_of_death = Dim2,
    death_rates = FactValueNumeric
  )

air_pollution_filtered_gbl

```  

#filtering out the rows with "Total" in them. 
```{r}
air_pollution_filtered_gbl <- air_pollution_filtered_gbl%>%
  filter(cause_of_death != "Total")
``` 


#cheching the type of death causes in the dataframe

```{r}
table(air_pollution_filtered_gbl['cause_of_death'])
```
```{r}
summary(air_pollution_filtered_gbl[ , c('death_rates')])
```


#Top 10 countries with the most deaths caused by air pollution.
```{r}
data_new2 <- air_pollution_filtered_gbl %>%                                     
  # Top N highest values by group
  arrange(desc(death_rates))%>% 
  group_by() %>%
  slice(1:10)
data_new2
```
 
```{r}
model_gbl <- lm(death_rates ~ continent, data= air_pollution_filtered_gbl)
summary(model_gbl)
```

```{r}
air_pollution_us 
```


