---
title: "Air polution"
output: html_notebook
---
```{r}
library(tidyverse)
library(sf)
library(ggplot2)
library(dplyr)
library(readxl)
library(hrbrthemes)
library(viridis)
library(forcats)
library(data.table)
library(scales)
library(leaflet)
```

#reading in the datasets that will be used in this project.

```{r}
countries_code <- read_csv("data/countries.csv.xls")
```
```{r}
world_countries <- read_csv("data/World_Countries_(Generalized).csv")
```
```{r}
world_countries <- world_countries %>%
  select(-c("FID", "ISO", "COUNTRYAFF","AFF_ISO"))
world_countries
```
```{r}
colnames(world_countries) <- c("country", "length", "area")
world_countries
```

```{r}
countries_code
```
```{r}
countries_code <- countries_code%>%
  select(-c("country"))
countries_code 
```

#Changing the column names so this dataframe can be merge to the others. 
```{r}
colnames(countries_code) <- c("latitude", "longitude", "country")
countries_code
```

```{r}
air_pollution_global <- read.csv("data/data.csv")
```

#the following datasets contain information of total population Nox and PM globally between 2003 - 2018

```{r}
population <- read_excel("data/aqdh-country-trends-major-air-pollutants-2003-2018.xlsx", sheet = "Population (GPWv4.11)" )
PM_Pop <- read_excel("data/aqdh-country-trends-major-air-pollutants-2003-2018.xlsx", sheet = "PM Pop.-Weighted (kg m^-3)")
```

#Droping duplicates in the dataframes and unnecesary columns. 

```{r}
population <- population %>% filter(!duplicated(.))
population = subset(population, select = -c(code))
```

```{r}
PM_Pop  <- PM_Pop %>% filter(!duplicated(.))
PM_Pop  = subset(PM_Pop, select = -c(code))
```

#Reshaping the dataframes into a large table.

```{r}
population <- population%>%
 pivot_longer(cols = c('2003', '2004', '2005', '2006', '2007', '2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018'),
               names_to = 'year',
               values_to = 'population')
```

```{r}
PM_Pop <- PM_Pop%>%
  pivot_longer(cols = c('2003', '2004', '2005', '2006', '2007', '2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018'),
               names_to = 'year',
               values_to = 'PM')
```

##This dataset corresponds to the overall air pollutants per country. 

#Merging the datasets into one dataframe. 
```{r}
all_population <- population %>%
  full_join(PM_Pop, by = c("iso","country", "year"))
```


Top 5 countries. 
```{r}
top_countries <- all_population %>%                                          
  arrange(desc(PM)) %>% 
  group_by(year) %>%
  slice(1:5)
top_countries
```
```{r}
ggplot(top_countries, aes(x=country, y=PM)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    scale_y_continuous(labels = scales::comma)+
    coord_flip() +
    xlab("") +
    theme_bw()
```

#PM particles per population
```{r}
all_population <- all_population %>% 
  mutate(pop_per100 = as.numeric(format(round(all_population$population/(all_population$PM/100000),1), nsmall=1)))

all_population
```


```{r}
merged_population <- merge(countries_code, all_population, by =  "country")
merged_population 
```

#linear model
```{r}
Pop_linear_model <- lm(pop_per100 ~ PM, data = all_population)
summary(Pop_linear_model)
```

#Log Linear model, and plot. 

```{r}
log_pop <- lm(pop_per100 ~ log(PM), data = all_population)
summary(log_pop)
```

**Global data comparing countries with the highest deaths caused by PM.** 

#Dropping extra information that is not needed.
```{r}
air_pollution_global
```
#Dropping duplicates 
```{r}
air_pollution_global <- air_pollution_global%>% filter(!duplicated(.))
```

```{r}
air_pollution_filtered_gbl <- subset(air_pollution_global, select = -c(IndicatorCode, ValueType, 
Location.type, Period.type, IsLatestYear, Dim1.type, Dim1ValueCode, Dim2.type, Dim2ValueCode, Dim3.type, Dim3, Dim3ValueCode, DataSourceDimValueCode, DataSource, FactValueNumericPrefix, FactValueUoM, FactValueNumericLowPrefix, FactValueNumericLow, FactValueNumericHighPrefix, FactValueNumericHigh, FactValueTranslationID, FactComments, Language, DateModified))
air_pollution_filtered_gbl
```
#Renaming columns in the Global Dataframe.
```{r}
air_pollution_filtered_gbl <- air_pollution_filtered_gbl%>%
  transmute(
    continent = ParentLocation,
    country = Location,
    sex = Dim1, 
    cause_of_death = Dim2,
    death_rates = FactValueNumeric
  )

air_pollution_filtered_gbl

```  

#filtering out the rows with "Total" in them. 
```{r}
air_pollution_filtered_gbl <- air_pollution_filtered_gbl%>%
  filter(cause_of_death != "Total")
``` 


#cheching the type of death causes in the dataframe
```{r}
air_pollution_filtered_gbl%>%
  count(cause_of_death, continent)%>%
  group_by()
  
```

#Top continents with the most deaths caused by air pollution.
```{r}
top_continent <- air_pollution_filtered_gbl %>%                                     
  # Top N highest values by group
  arrange(desc(death_rates))%>% 
  group_by(continent) %>%
  slice(1:5)
top_continent
```


```{r}

library(gapminder)

top_continent_plot <- top_continent %>% 
  arrange(desc(death_rates)) %>%
   group_by(continent) %>%
  ggplot(aes(x=continent, y=cause_of_death, size=death_rates, fill=continent)) +
  scale_size_continuous(labels = scales::comma) +
  theme(axis.text.x=element_text(color = "black", size=8, angle=30, vjust=.8, hjust=0.8))+
    geom_point(alpha=0.5, shape=21, color="black") +
    scale_size(range = c(.1, 15), name="Death Rates per Continent")+
   ylab("cause of death") +
    xlab("Death rates per continent") +
    theme(legend.position = "right")

top_continent_plot
```

```{r}
top_country_dr <- air_pollution_filtered_gbl %>%                                     
  # Top N highest values by group
  arrange(desc(death_rates))%>% 
  group_by() %>%
  slice(1:5)

top_country_dr
``` 
 #deaths per 100.000
```{r}

top_country_plot <- top_continent %>% 
  arrange(desc(death_rates)) %>%
   group_by(country) %>%
  ggplot(aes(x=cause_of_death, y=country, size=death_rates, fill=country)) +
  theme(axis.text.x=element_text(color = "black", size=10, angle=30, vjust=.9, hjust=0.8))+
    geom_point(alpha=0.5, shape=21, color="black") +
    scale_size(range = c(.1, 15), name="Death Rates per Continent")+
   ylab("cause of death") +
    xlab("Death rates per country") +
    theme(legend.position = "right",
          legend.title = element_text(colour = "black", size = 8, face = "bold"),
          legend.text = element_text(colour = "black", size = 8),
          legend.key = element_blank(),
          legend.key.size = unit(0.6, "cm"))
    
top_country_plot
``` 
```{r}
global_pop <- lm(death_rates ~ continent, data = air_pollution_filtered_gbl)
summary(global_pop)
```
```{r}
ggplot(global_pop, aes(x = continent, y = death_rates)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red")
```
#Attaching population data to death rates. 
```{r}
merged_global <- merge(air_pollution_filtered_gbl, merged_population , by = "country")
merged_global
```
```{r}
global_per100 <- merged_global%>% 
  mutate(death_per100 = as.numeric(format(round(merged_global$death_rates/(merged_global$population/100000),1), nsmall=1)))
global_per100
```
```{r}
death_sum = sum(merged_global$death_rates)
death_sum
```

```{r}
global_model<- lm(death_per100 ~ pop_per100, data = global_per100)
summary(global_pop)
```

#death rates as my target. Check the change over the years
```{r}
global_log <- lm(death_rates ~ PM, data = merged_global) 
summary(global_log)
```

#Merging datasets
```{r}
global <- merge(merged_global, world_countries, by = "country")
global
```


#converting to sf
```{r}
sf_global <- st_as_sf(global, coords = c("longitude", "latitude"), remove = FALSE)
sf_global
```

#Plotting a map
```{r}
bins = c(0,1,10,50,100,500)

pm_pal <- colorBin("Oranges", domain = global, bins = bins, na.color = "transparent")
plot_map <-  

pm_map <- leaflet() %>%
  addTiles()%>%
  addLayersControl(
      position = "bottomright",
      overlayGroups = c("PM", "death_rates"),
    options = layersControlOptions(collapsed = FALSE)) %>% 
   addProviderTiles(providers$CartoDB.Positron) %>%
  fitBounds(-100,-50,80,80) %>%
  addLegend("bottomright", pal = pm_pal, values = global$pop_per100,
            title = "<small>Particulate Matter per 100,000</small>")
  
pm_map

```




